What to simulate? Inferring the right direction for mental rotation

### 2014-02-04 man/cogsci-2014/cogsci-2014.tex
Minor changes here and there

% space between figure and caption
\addtolength{\abovecaptionskip}{-0.25cm}

% space below "Figure 1: ...", but only for inline figures
\addtolength{\textfloatsep}{-0.5cm}

@@

\title{What to simulate? Inferring the direction for mental rotation}

@@

\begin{abstract}
  When people use mental imagery to solve tasks, how do they decide
  \textit{which} images to generate? To answer this question, we
  explored how mental simulation should used in the classic
  psychological task of determining if two images depict the same
  object in different orientations \citeA{Shepard1971}. Through a
  rational analysis of mental rotation, we formalized four models and
  compared them to human performance on a 2D mental rotation
  task. None of the models were a particularly satisfying explanation
  of human behavior, leading us to suggest that the question of ``what
  to simulate?'' is more difficult than has previously been
  assumed. While only exploratory, this work provides the foundation
  necessary to begin answering this question.

@@

In the classic experiment by \citeA{Shepard1971}, participants viewed
images of three-dimensional objects and had to determine whether the
images depicted the same object (which differed by a rotation) or two
separate objects (which differed by a reflection and a rotation). The
result was that people's response times had a strong linear
correlation with the minimum angle of rotation, leading to the
conclusion that people solve this task by ``mentally rotating'' the
objects until they are congruent.  This conclusion leaves several
questions unanswered, however. How do people know the axis around
which to rotate the objects? If the axis is known, how do people know
which direction to rotate the objects?  And finally, how do people
know how long to rotate?

In this paper, we explore these questions through rational analysis
\cite{Marr:1983to,anderson90,Shepard:1987tt} and compare four models
of mental rotation. First, we discuss the previous literature on
mental imagery. Second, we outline computational- and
algorithmic-level analyses of the problem of mental rotation. Third,
we describe a behavioral experiment based on the classic mental
rotation studies \cite<e.g.,>{Cooper:1975wp}, and compare the results
of the experiment with each of the models. We conclude with a
discussion of the strengths and weaknesses of each model, and lay out
directions for future work.

\section{Background}

Previous models of mental rotation have, for the most part, focused on
the representation of mental images, rather than how people decide
\textit{which} mental images to generate. \citeA{Kosslyn:1977tv}
propose a model of the mental imagery buffer, but not the planning
process of how it is used. Similarly, \citeA{Julstrom:1985va} and
\citeA{Glasgow:1992tj} are mostly concerned with modeling the
representational format underlying imagery. Although
\citeA{Anderson1978} emphasizes the importance of considering both
representation and process, he dismisses the problem of determining
the direction of rotation as a ``technical difficulty''.

The only models (of which the authors are aware) that have seriously
attempted to address the decision of \textit{what} to simulate are
those by \citeA{Funt:1983wn} and \citeA{Just:1985uu}. In both of these
models, the axis and direction of rotation are computed prior to
performing the rotation. One object is then rotated through the target
rotation, and is checked against the other object for
congruency. However, this approach assumes that the corresponding
points on the two objects can be easily identified, which is not
necessarily the case.

@@

\subsection{A rational analysis of mental rotation}

@@

The original ``congruency'' hypothesis \cite{Shepard1971} actually is
a rational solution which achieves this goal, in the sense that the
smallest amount of computation coincides with rotating through the
minimum angle. However, it violates the constraint that we do not know
points correspondences, which is exactly what necessitates the use of
imagery. Noting that a rational solution need not necessarily maintain
a single trajectory of rotation, we explore an alternative rational
model, which--rather than computing the angle of rotation ahead of
time--engages in an ``active sampling'' strategy.

@@

@@

\subsection{Algorithmic assumptions}

@@

To determine the correct rotation, we solve for the rotation matrix by
computing $(\Xa \F_h^T)_\mathrm{left}^{-1}\cdot{}\Xb$, where
$(\Xa\F_h^T)_\mathrm{left}^{-1}$ is the left inverse of
$\Xa\F_h^T$. We then check each $h$ to see if the computation produces
a valid rotation matrix; the $h$ that does is the correct
hypothesis. This approach gives us the true value of $\theta$, so
Eq.~\ref{eq:lh-ratio} becomes a maximum likelihood ratio test, where
$\theta$ is set to the MLE value, rather than being marginalized:
\begin{equation}
  \ell = \frac{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hi)p(\hi)}{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hf)p(\hf)}.
  \label{eq:mle-lh-ratio}
\end{equation}

@@

\subsubsection{Hill climbing model}

@@

In the current formulation of the problem, choosing the threshold is
straightforward because we know both the exact geometry of the shapes
and that a linear transformation exists which will perfectly align
them. However, this choice is not always clear \textit{a priori}, as
the global optimum will depend on many factors (e.g., shape
complexity, dimensionality, perceptual uncertainty, and whether the
shapes are identical).  One way to deal with the problem of choosing a
threshold would simply be to perform an exhaustive search or use a
global optimization strategy. However, these strategies would not
result in the linear response time found by \citeA{Shepard1971}. A
second alternative, which could potentially demonstrate the linear
response time result, might instead just perform a Hill Climbing
(\Hc{}) search for a \textit{local} maximum. As with the \Oc{} and
\Th{} models, Eq.~\ref{eq:lh-ratio} becomes a maximum likelihood
ratio.

\subsubsection{Bayesian quadrature model}

@@

We randomly generated 20 shapes of 5 or 6 vertices (e.g.,
Fig.~\ref{fig:model-traces}). For each shape, we computed 20 ``same''
and 20 ``flipped'' stimuli pairs, with rotations ($\theta$) spaced at
$20^\prime$ increments between 0 and 360 (with 0 and 180 degree
rotations repeated twice, in order to gather an equal number of
responses for each angle between 0 and 180). ``Same'' pairs were
created by rotating $\Xa$ by $\theta$; ``flipped'' pairs were first
reflected $\Xa$ across the $y$-axis, then rotated by $\theta$.

We generated 5 additional shapes to be used in a practice block of 10
trials. Across these trials, each shape and each angle (60, 120, 180,
240, or 300) was repeated twice (once ``flipped'' and once ``same'')
such that no shape was presented at the same angle twice. We also
generated a sixth shape to be included along with the instructions.
This shape had both a ``flipped'' and ``same'' version, each rotated
to 320 degrees.

@@

All participants saw the same 10 practice trials, as described
above. There were 720 unique experimental stimuli (20 shapes $\times$
18 angles $\times$ 2 reflections), though because stimuli with
rotations of $0^\prime$ or $180^\prime$ were repeated twice, there
were 800 total experimental stimuli. These stimuli were split across 8
conditions in the following manner: first, stimuli were split into
four blocks of 200 trials. Within each block, each shape was repeated
10 times and each rotation was repeated 10 times (5 ``same'', 5
``flipped''), such that across all blocks, each stimulus appeared
exactly once. Each block was then split in half. Participants
completed two half-blocks (not necessarily both from the original full
block).

@@

Out of the 247 participants, 200 (81\%) were included in our
analyses. Of the rest, we excluded 10 (4\%) because of an experimental
error, 6 (2.4\%) because they had already completed a previous version
of the experiment, and 31 (12.6\%) because they failed a comprehension
check. A participant was marked as passing the comprehension check if
they answered at least 85\% of ``easy'' trials correctly. ``Easy''
trials were defined to be those for which the presented shapes
differed by a rotation of no more than 20 degrees. Additionally, we
excluded 82 trials for which the response time was either less than
100 milliseconds or greater than 20 seconds.

@@

There was a significant effect of learning both on response time
(\ExpTrialTimeCorr{}) and on accuracy (\ExpTrialAccuracyCorr{}),
though the effect on accuracy was not significant during the second
half of the experiment (\ExpaTrialAccuracyCorr{} for the first half
vs. \ExpbTrialAccuracyCorr{} for the second half). These practice
effects may have contributed to the not-quite-linearity of the human
response times; future work will need to collect more data per
participant to mitigate this effect.

@@

We set out to answer the question: how do people decide \textit{what}
to simulate when using mental imagery? Focusing on the specific case
of determining the direction and extent of mental rotation, we
formalized four models and compared their performance with the results
of a behavioral experiment. While it is clear that none of the models
fully captured human behavior on the mental rotation task, it is worth
noting where they shed light on aspects that have previously gone
unexplained.  For example, the \Bq{} model's adaptive stopping rule is
sensitive to the prior, and thus provides a possible explanation for
why people are slower to respond on ``flipped'' stimulus pairs.

Additionally, a closer look at individual shapes reveals several
interesting points (Fig.~\ref{fig:response-time-stimulus} shows curves
for one particular shape; these curves are fairly typical relative to
other shapes). First, the curves for the human response times on
specific shapes tend to either be linear, or linear first, then
flat. Second, while the \Th{} model appears to capture the concavity
of human response times overall, it is actually fairly inconsistent
within shapes. Third, the linearity of the \Bq{} model's response time
curves \textit{is} consistent within shapes, though it does not
exhibit the same ``plateau'' effect that sometimes appears in the
human curves.

>>>>

% space below "Figure 1: ...", but only for inline figures
%\addtolength{\textfloatsep}{-0.5cm}

@@

\title{What to simulate? Inferring the right direction for mental
  rotation}

@@

\begin{abstract}
  When people use mental imagery, how do they decide \textit{which}
  images to generate? To answer this question, we explored how mental
  simulation should used in the classic psychological task of
  determining if two images depict the same object in different
  orientations \cite{Shepard1971}. Through a rational analysis of
  mental rotation, we formalized four models and compared them to
  human performance. \TODO{None of the models were a particularly
    satisfying explanation of human behavior, leading us to suggest
    that the question of ``what to simulate?'' is more difficult than
    has previously been assumed.} While only exploratory, this work
  provides the foundation necessary to begin answering this question.

@@

Mental rotation provides a simple example of the decision problem
posed by simulation.  In the classic experiment by
\citeA{Shepard1971}, participants viewed images of three-dimensional
objects and had to determine whether the images depicted the same
object (which differed by a rotation) or two separate objects (which
differed by a reflection and a rotation). The result was that people's
response times had a strong linear correlation with the minimum angle
of rotation, leading to the conclusion that people solve this task by
``mentally rotating'' the objects until they are congruent.  This
conclusion leaves several questions unanswered, however. How do people
know the axis around which to rotate the objects? If the axis is
known, how do people know which direction to rotate the objects?  And
finally, how do people know how long to rotate?

In this paper, we explore these questions through rational analysis
\cite{Marr:1983to,anderson90,Shepard:1987tt} and compare four models
of mental rotation. We begin the paper by discussing the previous
literature on mental imagery. Next, we outline computational- and
algorithmic-level analyses of the problem of mental rotation.  We then
describe a behavioral experiment based on the classic mental rotation
studies \cite<e.g.,>{Cooper:1975wp}, and compare the results of the
experiment with each of the models. We conclude with a discussion of
the strengths and weaknesses of each model, and lay out directions for
future work.

Previous models of mental rotation have, for the most part, focused on
the representation of mental images, rather than how people decide
\textit{which} mental images to generate. \citeA{Kosslyn:1977tv}
proposed a model of the mental imagery buffer, but not the planning
process of how it is used. Similarly, \citeA{Julstrom:1985va} and
\citeA{Glasgow:1992tj} were mostly concerned with modeling the
representational format underlying imagery. Although
\citeA{Anderson1978} emphasized the importance of considering both
representation and process, he dismissed the problem of determining
the direction of rotation as a ``technical difficulty''.

The only models (of which the authors are aware) that seriously
attempted to address the decision of \textit{what} to simulate are
those by \citeA{Funt:1983wn} and \citeA{Just:1985uu}. In both of these
models, the axis and direction of rotation are computed prior to
performing the rotation. One object is then rotated through the target
rotation, and is checked against the other object for
congruency. However, this approach assumes that the corresponding
points on the two objects can be easily identified, which is not
necessarily the case.

@@

@@

The original ``congruency'' hypothesis \cite{Shepard1971} actually is
a rational solution to this problem, in the sense that the smallest
amount of computation coincides with rotating through the minimum
angle. However, it violates the constraint that we do not know the
points of correspondence between the images, which is exactly what
necessitates the use of imagery. Noting that a rational solution need
not necessarily maintain a single trajectory of rotation, we explore
an alternative rational model, which--rather than computing the angle
of rotation ahead of time--engages in an ``active sampling'' strategy.

@@

\section{How should we rotate?}

@@

\subsection{Algorithmic constraints}

@@

To determine the correct rotation, we solve for the rotation matrix by
computing $(\Xa \F_h^T)_\mathrm{left}^{-1}\cdot{}\Xb$, where
$(\Xa\F_h^T)_\mathrm{left}^{-1}$ is the left inverse of
$\Xa\F_h^T$. We then check each $h$ to see if the computation produces
a valid rotation matrix; the $h$ that does is the correct
hypothesis. This approach gives us the true value of $\theta$, so
Eq.~\ref{eq:lh-ratio} becomes a generalized likelihood ratio test,
where $\theta$ is set to the MLE value, rather than being
marginalized:
\begin{equation}
  \ell = \frac{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hi)p(\hi)}{\max_\theta p(\Xb\ \vert\ \Xa, \theta, \hf)p(\hf)}.
  \label{eq:mle-lh-ratio}
\end{equation}

@@

\subsubsection{Hill Climbing model}

@@

In the current formulation of the problem, choosing the threshold is
straightforward because we know both the exact geometry of the shapes
and that a linear transformation exists which will perfectly align
them. However, this choice is not always clear \textit{a priori}, as
the global optimum will depend on many factors (e.g., shape
complexity, dimensionality, perceptual uncertainty, and whether the
shapes are identical).  One way to deal with the problem of choosing a
threshold would simply be to perform an exhaustive search or use a
global optimization strategy. However, these strategies would not
result in the linear response time found by \citeA{Shepard1971}. A
second alternative, which could potentially demonstrate the linear
response time result, might instead just perform a Hill Climbing
(\Hc{}) search for a \textit{local} maximum. As with the \Oc{} and
\Th{} models, Eq.~\ref{eq:lh-ratio} becomes a generalized likelihood
ratio test.

\subsubsection{Bayesian Quadrature model}

@@

We randomly generated 20 shapes of five or six vertices (e.g.,
Fig.~\ref{fig:model-traces}). For each shape, we computed 20 ``same''
and 20 ``flipped'' stimuli pairs, with rotations ($\theta$) spaced at
$20^\prime$ increments between 0 and 360 (with 0 and 180 degree
rotations repeated twice, in order to gather an equal number of
responses for each angle between 0 and 180). ``Same'' pairs were
created by rotating $\Xa$ by $\theta$; ``flipped'' pairs were first
reflected $\Xa$ across the $y$-axis, then rotated by $\theta$.

We generated five additional shapes to be used in a practice block of
10 trials. Across these trials, each shape and each angle (60, 120,
180, 240, or 300) was repeated twice (once ``flipped'' and once
``same'') such that no shape was presented at the same angle twice. We
also generated a sixth shape to be included along with the
instructions.  This shape had both a ``flipped'' and ``same'' version,
each rotated to 320 degrees.

@@

All participants saw the same 10 practice trials, as described
above. There were 720 unique experimental stimuli (20 shapes $\times$
18 angles $\times$ 2 reflections), though because stimuli with
rotations of $0^\prime$ or $180^\prime$ were repeated twice, there
were 800 total experimental stimuli. These stimuli were split across
eight conditions in the following manner: first, stimuli were split
into four blocks of 200 trials. Within each block, each shape was
repeated ten times and each rotation was repeated ten times (five
``same'', five ``flipped''), such that across all blocks, each
stimulus appeared exactly once. Each block was then split in
half. Participants completed two half-blocks (not necessarily both
from the original full block).

@@

Out of the 247 participants, 200 (81\%) were included in our
analyses. Of the rest, we excluded 10 (4\%) because of an experimental
error, 6 (2.4\%) because they had already completed a related
experiment, and 31 (12.6\%) because they failed a comprehension
check. A participant was marked as passing the comprehension check if
they answered at least 85\% of ``easy'' trials correctly. ``Easy''
trials were defined to be those for which the presented shapes
differed by a rotation of no more than 20 degrees. Additionally, we
excluded 82 trials for which the response time was either less than
100 milliseconds or greater than 20 seconds.

@@

There was a significant effect of trial number both on response time
(\ExpTrialTimeCorr{}) and on accuracy (\ExpTrialAccuracyCorr{}),
though the effect on accuracy was not significant during the second
half of the experiment (\ExpaTrialAccuracyCorr{} for the first half
vs. \ExpbTrialAccuracyCorr{} for the second half). These practice
effects may have contributed to the not-quite-linearity of the human
response times; future work will need to collect more data per
participant to mitigate this effect.

@@

We set out to answer the question of how people decide \textit{what}
to simulate when using mental imagery. Focusing on the specific case
of determining the direction and extent of mental rotation, we
formalized four models and compared their performance with the results
of a behavioral experiment. While it is clear that none of the models
fully captured human behavior on the mental rotation task, it is worth
noting where they shed light on aspects that have previously gone
unexplained.  For example, the \Bq{} model's adaptive stopping rule is
sensitive to the prior, and thus provides a possible explanation for
why people are slower to respond on ``flipped'' stimulus pairs.

Additionally, a closer look at performance on individual shapes
reveals several interesting points
(Fig.~\ref{fig:response-time-stimulus} shows curves for one particular
shape; these curves are fairly typical relative to other
shapes). First, the curves for the human response times on specific
shapes tend to either be linear, or linear first, then flat. Second,
while the \Th{} model appears to capture the concavity of human
response times overall, it is actually fairly inconsistent within
shapes. Third, the linearity of the \Bq{} model's response time curves
\textit{is} consistent within shapes, though it does not exhibit the
same ``plateau'' effect that sometimes appears in the human curves.
